Efficiently stable presentations from error-correcting codes

### 2023-10-24 efficient_stability.tex
Updates from Overleaf

\subsection{The $L^\infty$ analogue}
As discussed in the introduction (see Remark \ref{rem:L^infty_analogue_defn}), it is more common in stability literature to use a $L^\infty$ analogue of Definition~\ref{def:eff-stab}, where the notion of almost-homomorphism and closeness are both measured by taking a supremum over relations and generators respectively, as opposed to averaging according to distributions $\mu_R,\mu_S$. 
A slight modification of the example given in  this appendix to prove Lemma~\ref{lem:lower_bound_on_stability_rate_standard_presentation_Z_2^k} provides a lower bound also in that setup. The idea is actually quite general, so we frame it in a somewhat general fashion.

@@

  Then, $\sigma$ extends (by the universal property of the free group) to an automorphism of $\mF(S)$ which we still denote by $\sigma$. 

@@

    We denote by $\mM_\Phi=\bigoplus_{\alpha\in \Phi}\mM\subseteq \mM_\infty$ and by $\mM_\alpha$ the copy of $\mM$ at the $\alpha\in \Phi$ position.\tnote{Is it canonical what is the trace on $\mM_\Phi$? In the calculation below you assume a uniform weighted average, is it useful to say it?}
    Let $w_i=\mu_R(R_i)$, namely the probability that $\mu_R$ samples a relation from the orbit $R_i$. Then, for every $r\in R_i$, we have \tnote{For the second line, is it clear that we get uniform over $R_i$?}

@@

        w_i\|\rho'(r)-\Id\|_\tau^2&=w_i\Expectation_{\alpha\in \Phi} \|\rho(\alpha(r))-\Id\|_\tau^2\\
        &=w_i\Expectation_{r'\in R_i} \|\rho(r')-\Id\|_\tau^2\\
        &\leq \sum_j w_j\Expectation_{r'\in R_j} \|\rho(r')-\Id\|_\tau^2\\
        &=\Expectation_{r\sim \mu_R}\|\rho(r)-\Id\|_\tau^2\leq \eps.

@@

    \forall r\in R\ \colon\ \|\rho'(r)-\Id\|_\tau^2\leq \max({\nicefrac{1}{w_i}})\cdot\eps,

@@

which in turn means that $\rho'$ is a $\max_i\{{\nicefrac{1}{w_i}}\}\cdot\eps$-approximate representation in the $L^\infty$ sense. 

@@

    By applying this construction on the example from Claim~\ref{claim:appendix1}, the resulting $\rho'$ is a $\nicefrac{2|E|}{\binom{k}{2}}$-approximate representation of \eqref{eq:z2-efficient} in the $L^\infty$ sense.\tnote{Can we expand on this? Any permutation is an automorphism, right? So we end up having two classes of relations, and $w_1=w_2=1/2$. It might just be worth spelling out}

@@

Now, assume that for any genuine representation $\varphi\colon \Gamma\to \mU(\mN)$, where $\mN=P\mM_\infty P$, and every isometry $w\in P\mM_\infty \Id_\mM$, we have 
\max\left\{\Expectation_{s\sim \mu_S} \|\rho(s)-w^*\varphi(s)w\|_\tau^2\ ,\ \tau^\mM(\Id_\mM-w^*w)\right\}\geq \delta.
Let $\varphi'\colon \Gamma\to \mU(\mN)$ be a genuine representation of $\Gamma$ satisfying \tnote{Is the $1/|\Phi|$ in front correct? Or maybe I'm missing how the $\tau_\infty$ is normalized exactly. It just sums over the $\alpha$'s?}\tnote{Also, I'm not sure I understood your starting point. You want to assume that $\varphi'$ is $L_\infty$-close, right? Here you only do $L_1$. Is this what the last sentence of the proof is supposed to justify? If so, perhaps better epxlain here.}
\frac{1}{|\Phi|}\Expectation_{s\sim \mu_S} \|\rho'(s)-w^*\varphi'(s)w\|_\tau^2\ ,\ \tau^{\mM_\Phi}(\Id_{\mM_\Phi}-w^*w)\leq  \delta',
where we again abuse notation and think of $\|X\|_\tau^2=\tau_\infty(X^*X)$.
As before, we can assume $w=PI_{\mM_\Phi}$. Then,
    \Expectation_{\alpha\in \Phi}\Expectation_{s\sim \mu_S} \|\rho(\alpha(s))-\Id_{\mM_\alpha}\varphi'(s)\Id_{\mM_\alpha}\|_\tau^2&=\frac{1}{|\Phi|}\Expectation_{s\sim \mu_S}\sum_{\alpha\in \Phi}\|\Id_{\mM_\alpha}\rho'(s)\Id_{\mM_\alpha}-\Id_{\mM_\alpha}\varphi'(s)\Id_{\mM_\alpha}\|_\tau^2 \\
    &\leq \frac{1}{|\Phi|}\Expectation_{s\sim \mu_S}\|\rho'(s)-\Id_{\mM_\Phi}\varphi'(s)\Id_{\mM_\Phi}\|_\tau^2\leq \delta'.
In particular, by Markov's inequality, for at least two thirds of the $\alpha\in \Phi$ we have 
\Expectation_{s\sim \mu_S} \|\rho(\alpha(s))-\Id_{\mM_\alpha}\varphi'(s)\Id_{\mM_\alpha}\|_\tau^2\leq 3\delta'.
Similarly,
    \delta'&\geq \tau^{\mM_\Phi}(\Id_{\mM_\Phi}-\Id_{\mM_\Phi}P\Id_{\mM_\Phi})\\
    &=\Expectation_{\alpha\in\Phi}\tau_\infty(\Id_{\mM_\alpha}-\Id_{\mM_\alpha}P\Id_{\mM_\alpha})
\end{split}\;,
and for at least two thirds of the $\alpha\in \Phi$ we have 
\tau^{\mM_\alpha}(\Id_{\mM_\alpha}-\Id_{\mM_\alpha}P\Id_{\mM_\alpha})\leq 3\delta'.
Hence, $\delta'\geq \frac{\delta}{3}$.
Since the $L^\infty$ distance between functions is larger than the $L^1$ distance between them (with respect to any fully supported measure $\mu_S$), we conclude that a lower bound of $f(\eps)$ on the modulus of stability of a sufficiently symmetric presentation induces a lower bound of $\frac{\min{w_i}}{3}f(\eps)$ \tnote{on? (I know, but maybe we should say?)}.

\bibliography{qld}

>>>>

\subsection{The $L^\infty$ analogue}
As discussed in the introduction (see Remark \ref{rem:L^infty_analogue_defn}), it is more common in stability literature to use a $L^\infty$ analogue of Definition~\ref{def:eff-stab}, where the notion of almost-homomorphism and closeness are both measured by taking a supremum over relations and generators respectively, as opposed to averaging according to distributions $\mu_R,\mu_S$. 
Let us recall the exact definition. We say that a homomorphism $\rho\colon \mF(S)\to \mU(\mM)$ is an $(\eps,\infty)$-approximate representation if 
\[
\forall r\in R\ \colon \ \ \|\rho(r)-\Id_\mM\|_\tau^2\leq \eps.
\]
Furthermore, homomorphisms $\rho\colon \mF(S)\to \mU(\mM),\varphi\colon \mF(S)\to \mU(\mN)$ are $(\delta,\infty)$-close if there exists an isometry $w\in P\mM_\infty \Id_\mM$ such that
\[
\forall s\in S\colon\ \ \|\rho(s)-w^*\varphi(s)w\|_\tau^2\leq \delta.
\]
The goal of this subsection of the appendix is to provide a somewhat general procedure to convert lower bounds on the modulus of stability with respect to $\mu_S,\mu_R$ into a lower bound on the $L^\infty$ modulus of stability.

@@

  Then, $\sigma$ extends (by the universal property of the free group) to an automorphism of $\mF(S)$ which we still denote by $\sigma$ as well. 

@@

    We denote by $\mM_\Phi=\bigoplus_{\alpha\in \Phi}\mM\subseteq \mM_\infty$ and by $\mM_\alpha$ the copy of $\mM$ at the $\alpha\in \Phi$ position.\tnote{Is it canonical what is the trace on $\mM_\Phi$? In the calculation below you assume a uniform weighted average, is it useful to say it?}\mnote{Adding a clarification:} Note that $\mM_\Phi$ embeds in $\mM_\infty$, e.g. in the coordinates $1,...,|\Phi|$, and inherits a trace from it by defining 
    \tau^{\mM_\Phi}(\bigoplus A_\alpha)=\frac{\tau_\infty(\bigoplus A_\alpha)}{\tau_\infty(\Id_{\mM_\Phi})}=\frac{1}{|\Phi|}\sum_{\alpha\in \Phi}\tau^\mM(A_\alpha)=\Expectation_{\alpha\in \Phi}\tau^\mM(A_\alpha).
    To be consistent, whenever we use $\|X\|_\tau^2$ it means $$\tau_\infty(X^*X)=\tau_\mM(X^*X)=|\Phi|\cdot\tau_{\mM_\Phi}(X^*X).$$
    Let $w_i=\mu_R(R_i)$, namely the probability that $\mu_R$ samples a relation from the orbit $R_i$. Since we assumed $\mu_R$ is uniform over orbits, and since $\Phi$ acts transitively on each orbit, we know that 
    \forall r\in R_i\ \colon \ \ \Expectation_{\alpha\in \Phi}f(\alpha(r))=\Expectation_{r'\in R_i}f(r)
    for any function $f\colon \mF(S)\to \complex$. Therefore, \tnote{For the second line, is it clear that we get uniform over $R_i$?}

@@

       \forall r\in R_i\ \colon \ \  \frac{w_i}{|\Phi|}\|\rho'(r)-\Id_{\mM_\Phi}\|_\tau^2&=w_i\Expectation_{\alpha\in \Phi} \|\rho(\alpha(r))-\Id_\mM\|_\tau^2\\
        &=w_i\Expectation_{r'\in R_i} \|\rho(r')-\Id_\mM\|_\tau^2\\
        &\leq \sum_j w_j\Expectation_{r'\in R_j} \|\rho(r')-\Id_\mM\|_\tau^2\\
        &=\Expectation_{r\sim \mu_R}\|\rho(r)-\Id_\mM\|_\tau^2\leq \eps.

@@

    \forall r\in R\ \colon\ \frac{1}{|\Phi|}\|\rho'(r)-\Id_{\mM_\Phi}\|_\tau^2\leq \max({\nicefrac{1}{w_i}})\cdot\eps,

@@

which in turn means that $\rho'$ is a $(\max_i\{{\nicefrac{1}{w_i}}\}\cdot\eps,\infty)$-approximate representation --- since $$\frac{1}{|\Phi|}\|\cdot\|_\tau^2=\|\cdot\|_{\tau^{\mM_\Phi}}^2,$$
which is the relevant parameter to consider.
\begin{corollary}
    Any $(\eps,\mu_R)$-approximate representation of $\langle S\colon R\rangle$, where $\mu_R$ is uniform over orbits of $\Phi=\textrm{Aut}(\langle S\colon R\rangle)$, can be transformed into a $(\max_i\{{\nicefrac{1}{w_i}}\}\cdot\eps,\infty)$-approximate representation. 
\end{corollary}

@@

    By applying this construction on the example from Claim~\ref{claim:appendix1}, the resulting $\rho'$ is a $(\nicefrac{2|E|}{\binom{k}{2}},\infty)$-approximate representation of \eqref{eq:z2-efficient}.\tnote{Can we expand on this? Any permutation is an automorphism, right? So we end up having two classes of relations, and $w_1=w_2=1/2$. It might just be worth spelling out}\mnote{Done:} This is because $\Phi=\textrm{Sym}(S)$, and there are two  orbits --- the commutation relations and the involutions --- where $\mu_R$ is supprted equally on each of them. Namely, $w_{commutation}=w_{involution}=\nicefrac{1}{2}$.

@@

Recall that our goal is to translate $L^1$ lower bounds into $L^\infty$ lower bounds. Namely, we would like to deduce that, if every genuine representation of $\Gamma$ is $(\delta,\mu_S)$-far from $\rho$, then every genuine representation of $\Gamma$ is $(\delta',\infty)$-far from $\rho'$, for $\delta'=C\delta$.  Since $\mu_S$ was assumed to be fully supported, the $L^\infty$ distance is lower bounded by the $L^1$-distance, and being $(\delta',\mu_S)$-far from $\rho'$ implies being $(\delta',\infty)$-far from it. Thus, we can forget about the $L^\infty$ notion of distance and lower bound our usual notion of distance.

To that end, assume that every genuine representation of $\Gamma$ is $(\delta,\mu_S)$-far from $\rho$. Namely, for every genuine representation $\varphi\colon \Gamma\to \mU(\mN)$, where $\mN=P\mM_\infty P$, and every isometry $w\in P\mM_\infty \Id_\mM$, we have 
\[
\max\left\{\Expectation_{s\sim \mu_S} \|\rho(s)-w^*\varphi(s)w\|_\tau^2\ ,\ \tau^\mM(\Id_\mM-w^*w)\right\}\geq \delta.
\]
Let $\varphi'\colon \Gamma\to \mU(\mN)$ be a genuine representation of $\Gamma$ which is $(\delta',\mu_S)$-close to $\rho'$, i.e.\ it satisfies \tnote{Is the $1/|\Phi|$ in front correct? Or maybe I'm missing how the $\tau_\infty$ is normalized exactly. It just sums over the $\alpha$'s?}\tnote{Also, I'm not sure I understood your starting point. You want to assume that $\varphi'$ is $L_\infty$-close, right? Here you only do $L_1$. Is this what the last sentence of the proof is supposed to justify? If so, perhaps better epxlain here.}\mnote{I added an explanation before the analysis, immediately after Remark A.5.}
\[
\frac{1}{|\Phi|}\Expectation_{s\sim \mu_S} \|\rho'(s)-w^*\varphi'(s)w\|_\tau^2\ ,\ \tau^{\mM_\Phi}(\Id_{\mM_\Phi}-w^*w)\leq  \delta',
\]
where  again  $\|X\|_\tau^2=\tau_\infty(X^*X)$.
As before, we can assume $w=PI_{\mM_\Phi}$. Then,
\[
\begin{split}
    \Expectation_{\alpha\in \Phi}\Expectation_{s\sim \mu_S} \|\rho(\alpha(s))-\Id_{\mM_\alpha}\varphi'(s)\Id_{\mM_\alpha}\|_\tau^2&=\frac{1}{|\Phi|}\Expectation_{s\sim \mu_S}\sum_{\alpha\in \Phi}\|\Id_{\mM_\alpha}\rho'(s)\Id_{\mM_\alpha}-\Id_{\mM_\alpha}\varphi'(s)\Id_{\mM_\alpha}\|_\tau^2 \\
    &\leq \frac{1}{|\Phi|}\Expectation_{s\sim \mu_S}\|\rho'(s)-\Id_{\mM_\Phi}\varphi'(s)\Id_{\mM_\Phi}\|_\tau^2\leq \delta'.
\end{split}
\]
In particular, by Markov's inequality, for at least two thirds of the $\alpha\in \Phi$ we have 
\[
\Expectation_{s\sim \mu_S} \|\rho(\alpha(s))-\Id_{\mM_\alpha}\varphi'(s)\Id_{\mM_\alpha}\|_\tau^2\leq 3\delta'.
\]
Similarly,
\[
\begin{split}
    \delta'&\geq \tau^{\mM_\Phi}(\Id_{\mM_\Phi}-\Id_{\mM_\Phi}P\Id_{\mM_\Phi})\\
    &=\Expectation_{\alpha\in\Phi}\tau_\infty(\Id_{\mM_\alpha}-\Id_{\mM_\alpha}P\Id_{\mM_\alpha})
\end{split}\;,
\]
and for at least two thirds of the $\alpha\in \Phi$ we have 
\[
\tau^{\mM_\alpha}(\Id_{\mM_\alpha}-\Id_{\mM_\alpha}P\Id_{\mM_\alpha})\leq 3\delta'.
\]
Hence, $\delta'\geq \frac{\delta}{3}$, and we deduce that every genuine representation of $\Gamma\cong \langle S\colon R\rangle$ is at least $(\nicefrac{\delta}{3},\infty)$-away from $\rho'$.
%Since the $L^\infty$ distance between functions is larger than the $L^1$ distance between them (with respect to any fully supported measure $\mu_S$), we conclude that a lower bound of $f(\eps)$ on the modulus of stability of a sufficiently symmetric presentation induces a lower bound of $\frac{\min{w_i}}{3}f(\eps)$ 
\tnote{on? (I know, but maybe we should say?)}.
\begin{corollary}
    The transformation we described  $\rho\mapsto\rho'$ translates a $(\eps,\mu_R)$-approximate representation into a $((\max{\nicefrac{1}{w_i}})\cdot\eps,\infty)$-approximate representaion, and if every genuine representation of $\Gamma$ is $(\delta,\mu_S)$-far from $\rho$, then every genuine representation is $(\nicefrac{\delta}{3},\infty)$-far from $\rho'$.
\end{corollary}
\begin{remark}
    By applying this corollary to the construction from the beginning of the appendix, we conclude that the $L^\infty$ modulus of stability of the presentation \eqref{eq:z2-efficient} is $\Omega(k\eps)$.
\end{remark}
\bibliography{qld}
